Resolvable Ambiguity

### 2019-02-25 paper/paper.tex
Description of the basic algorithm

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan]{acmart}\settopmatter{}

@@

@@

@@

%% symbols used throughout the paper

\newcommand{\NT}{\mathbb{N}} % Set of nonterminals
\newcommand{\T}{\mathbb{T}} % Set of terminals

@@

% TODO: ref for undecidable ambiguity checking
To prevent the possibility of an end-user encountering such an error we must ensure that the grammar cannot give rise to an unresolvable ambiguity. It is worth mentioning here that statically checking if a context-free grammar is ambiguous has long been known to be undecidable. Unresolvable ambiguity, however, turns out to be decidable\footnote{cross your fingers}.

@@

\begin{itemize}
  % TODO: reference
  \item Building on PADL-PAPER, which merely isolates ambiguities, an algorithm that suggests solutions to ambiguity errors.
  \item A formalization of the unresolvable ambiguity property for context-free EBNF grammars.
  \item An algorithm for deciding if a grammar is unresolvably ambiguous or not.
\end{itemize}

@@

% TODO: the last line feels a bit odd, notationally
\begin{figure}
  \begin{tabular}{@{}ll@{}}
      Terminals & $t \in \T$ \\
      Non-terminals & $N \in \NT$ \\
      Regular expressions & $r ::= t \mid N \mid r \cdot r \mid r + r \mid \epsilon \mid r^{*}$ \\
      Productions & $N -> r$ \\
  \end{tabular}
  \caption{Context-free EBNF grammars}
  \label{fig:grammar-definition}
\end{figure}

A context-free EBNF grammar is a tuple $(S, P, \NT, \T)$. $S \in \NT$ is the starting symbol, $P$ is a set of productions, as given in Figure~\ref{fig:grammar-definition}, $\NT$ and $\T$ are disjoint sets of non-terminals and terminals, respectively.

The (word) language of a given non-terminal $N$ in a grammar $G = (S, P, \NT, \T)$ is given by $L_G(N)$:

@@

The word language of a grammar $G$, written $L(G)$, is thus $L_G(S)$. We will omit the subscript whenever the intended grammar is clear from the context.

@@

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

%% Bibliography
%\bibliography{bibfile}

%% Appendix
\appendix
\section{Appendix}

Text of appendix \ldots

>>>>

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan]{acmart}\settopmatter{}

@@

\usepackage{tikz}
\usetikzlibrary{arrows,automata}
\usepackage{tikz-qtree} % Used for syntax trees
\usepackage{tikz-cd} % Used for commutative diagrams

@@

\usepackage{marginnote}

@@

%% symbol definitions used throughout the paper

\newcommand{\NT}{\mathbb{N}} % Set of nonterminals
\newcommand{\T}{\mathbb{T}} % Set of terminals
\newcommand{\I}{\mathbb{I}} % Set of identifiers
\newcommand{\yield}{\mathit{yield}} % yield of a parse tree
\newcommand{\semantic}{\mathit{semantic}} % remove semantically unimportant productions from a w' \in L(G')
\newcommand{\parse}{\mathit{parse}} % go from a w' \in L(G') to a subset of L(T)

@@

% TODO: ref for undecidable ambiguity checking
To prevent the possibility of an end-user encountering such an error we must ensure that the grammar cannot give rise to an unresolvable ambiguity. It is worth mentioning here that statically checking if a context-free grammar is ambiguous has long been known to be undecidable. Unresolvable ambiguity, however, turns out to be decidable\footnote{With some caveats, I'll talk more about this during the meeting.}.

@@

\begin{itemize}
  \item Building on \cite{palmkvistCreatingDomainSpecificLanguages2019}, which merely isolates ambiguities, an algorithm that suggests solutions to ambiguity errors.
  \item A formalization of the unresolvable ambiguity property for context-free EBNF grammars.
  \item An algorithm for deciding if a grammar is unresolvably ambiguous or not.
\end{itemize}

@@

\begin{quote} % TODO: remove
  \textbf{NOTE:} The text hereafter is written a bit later than the text before, and is ever so slightly inconsistent with it.
\end{quote}

% TODO: the last line feels a bit odd, notationally
\begin{figure}
  \begin{tabular}{@{}ll@{}}
      Terminals & $t \in \T$ \\
      Non-terminals & $N \in \NT$ \\
      Identifiers & $i \in \I$ \\
      Regular expressions & $r ::= t \mid N \mid r \cdot r \mid r + r \mid \epsilon \mid r^{*}$ \\
      Productions & $i : N -> r$ \\
  \end{tabular}
  \caption{Context-free EBNF grammars}
  \label{fig:grammar-definition}
\end{figure}

A context-free EBNF grammar is a tuple $(S, P, \NT, \T, \I)$. $S \in \NT$ is the starting symbol, $P$ is a set of productions, as given in Figure~\ref{fig:grammar-definition}, $\NT$, $\T$, and $\I$ are sets of non-terminals, terminals, and identifiers, respectively. Additionally, we require that $\NT \cap \T = \emptyset$, and that the identifiers uniquely identify each production, i.e., there are no two distinct productions in $P$ with the same identifier.

The (word) language of a given non-terminal $N$ in a grammar $G = (S, P, \NT, \T, \I)$ is given by $L_G(N)$:

@@

\noindent The word language of a grammar $G$, written $L(G)$, is thus $L_G(S)$. We will omit the subscript whenever the intended grammar is clear from context.

Note that the right hand side of a production is a regular expression, which could potentially be ambiguous in and of itself. For this work we assume the rhs regular expressions to either be unambigous, or that any remaining ambiguity is unimportant. This can be achieved in a number of ways, for example using the ambiguity checking of Brabrand and Thomsen \cite{brabrandTypedUnambiguousPattern2010}.

As such, for a given grammar $G = (S, P, \NT, \T, \I)$ we construct a linear representation of its parse trees as another grammar $T_G = (S, P', \NT, \T', \I)$ where:

$$
\begin{array}{r@{\;=\;}l}
  P' & \{i : N -> [_i r ]_i \mid (i : N -> r) \in P \} \\
  \T' & \T \sqcup \bigcup_{(i : N -> r) \in P} \{[_i, ]_i \} \\
\end{array}
$$

\noindent where $\sqcup$ denotes disjoint union. Intuitively, we surround the right hand side of each production with a unique pair of brackets, signifying the production that was used for the parse. Again, we will omit the subscript whenever the intended grammar is clear from context.

The $\yield$ of a parse tree is the word it parsed, i.e., $\yield : L(T) -> L(G)$. Intuitively, $\yield$ removes the brackets introduced when constructing $T$.

A given word $w \in L(G)$ is ambiguous iff:

$$\exists t_1, t_2 \in L(T).\ \yield(t_1) = w \land \yield(t_2) = w \land t_1 \neq t_2$$

\noindent A grammar is ambiguous iff it contains at least one ambiguous word.

\section{Parse-time Disambiguation}

We begin this section with some motivation, and then list our definition of \emph{resolvable ambiguity}.

We can divide the productions present in a programming language grammar in two groups: those that are semantically important, and those that are semantically \emph{un}important. The former group covers most productions, while the latter contains, e.g., parentheses used for explicit grouping. If programs were written directly as syntax trees then the latter would be unnecessary; two syntax trees that differ only by parentheses are semantically the same.

Thus we wish the output of parsing to be a syntax tree consisting entirely of semantically important productions. This distinction is useful to make, because it allows us to decouple \emph{what} we want to be expressible and \emph{how} it is to be expressed. As an example, in Figure~\ref{fig:example-grammar}, \ref{fig:example-grammar:ambig-grammar} is the \emph{what} and \ref{fig:example-grammar:unambig-grammar} is the \emph{how}. The latter grammar is a modification of the former that adds precedence, associativity, and parentheses, yielding an unambiguous grammar with at least one way to express each semantically distinct tree.

\begin{figure}
  \begin{subfigure}{\linewidth}
    \setlength{\grammarindent}{5em}
    \begin{grammar}
      <expr> ::= Sum: <expr> '+' <expr>
        \alt Product: <expr> '*' <expr>
        \alt Number: <number>
    \end{grammar}
    \caption{The (ambiguous) intuitive grammar without parentheses.}
    \label{fig:example-grammar:ambig-grammar}
  \end{subfigure}

  \begin{subfigure}{\linewidth}
    \setlength{\grammarindent}{5em}
    \begin{grammar}
      <expr> ::= Sum: <product> '+' <expr>
        \alt PassthroughE: <product>

      <product> ::= Product: <atom> '*' <product>
        \alt PassthroughP: <atom>

      <atom> ::= Paren: '(' <expr> ')'
        \alt Number: <number>
    \end{grammar}
    \caption{The (unambiguous) grammar with parentheses.}
    \label{fig:example-grammar:unambig-grammar}
  \end{subfigure}

  \begin{subfigure}{\linewidth}
    \begin{center}
    \Tree [.PassthroughE
      [.\underline{Product}
        [.Paren
          '('
          [.\underline{Sum}
            [.PassthroughP [.\underline{Number} 1 ] ]
            '+'
            [.PassthroughE [.PassthroughP [.\underline{Number} 2 ] ] ] ]
          ')' ]
        '*'
        [.PassthroughP [.\underline{Number} 3 ] ] ] ]
    \end{center}
    \caption{Parse tree for $(1 + 2) * 3$. The underlined nodes are semantically important.}
    \label{fig:example-grammar:tree}
  \end{subfigure}

  \begin{subfigure}{\linewidth}
    \begin{center}
    \Tree [.\underline{Product}
      [.\underline{Sum} [.\underline{Number} 1 ] '+' [.\underline{Number} 2 ] ]
      '*'
      [.\underline{Number} 3 ] ]
    \end{center}
    \caption{The same parse tree, after removing the semantically unimportant nodes.}
    \label{fig:example-grammar:sem-tree}
  \end{subfigure}

  \caption{A basic expression grammar in two variations, and two example syntax trees.}
  \label{fig:example-grammar}
\end{figure}

Our definition thus refers to four grammars in total:

\begin{itemize}
  \item The semantic grammar $G$, containing only the semantically important productions (e.g., Figure~\ref{fig:example-grammar:ambig-grammar}).
  \item $T_G$ (generally abbreviated as $T$), the parse trees of $G$, representing the trees that must be expressible (Figure~\ref{fig:example-grammar:sem-tree} is in this language).
  \item The parse grammar $G'$, a modification of $G$ meant to actually be used for parsing (e.g., Figure~\ref{fig:example-grammar:unambig-grammar}). The details of these modifications will be covered later. % TODO: likely put a forward reference here
  \item $T_{G'}$ (generally abbreviated as $T'$), the parse trees of $G'$ (Figure~\ref{fig:example-grammar:tree} is in this language).
\end{itemize}

\noindent We also require a function $\semantic : L(T') -> L(T)$ that removes the semantically unimportant productions from a parse tree. The relation between the four grammars can be seen in Figure~\ref{fig:grammar-square}. Finally, we define the function $\parse : L(G') -> 2^{L(T)}$:

$$
\parse(w') = \{ \semantic(t') \mid t' \in L(T') \land \yield(t') = w' \}
$$

\begin{figure}
  \begin{tikzcd}
    L(G) & L(T) \arrow[l, "\yield"] \\
    L(G') & L(T') \arrow[u, "\semantic"'] \arrow[l, "\yield"]
  \end{tikzcd}
  \caption{The grammars considered, and their relation to each other. $G$ is provided by a user of the system, along with instructions how to modify $G$ to construct $G'$, while $T$ and $T'$ are automatically derived.}
  \label{fig:grammar-square}
\end{figure}

%% \begin{definition}
%%   A language defined by the semantic grammar $G$ and the parse grammar $G'$ is resolvable if:
%%   $$
%%   \begin{array}{l}
%%   \forall t \in L(T).\\
%%   \quad \exists w' \in L(G').\\
%%   \qquad \forall t' \in L(T'). \yield(t') = w' -> \semantic(t') = t
%%   \end{array}
%%   $$
%% \end{definition}

\begin{definition}
  A language defined by the semantic grammar $G$ and the parse grammar $G'$ is resolvable if:
  $$
  \begin{array}{l}
  \forall t \in L(T).\\
  \quad \exists w' \in L(G').\ \parse(w') = \{t\}
  \end{array}
  $$
\end{definition}

\noindent Intuitively, a language is resolvably ambiguous if there is at least one (unambiguous\footnote{For a slightly different meaning of unambiguous, it's here relating $G'$ to $T$, instead of $G'$ to $T'$.}) word for each semantically distinct tree. Note that neither $G$ nor $G'$ necessarily need to be unambiguous for this to hold.

\section{Static Resolvability Check}

This section describes a decision algorithm for detecting unresolvable ambiguities, starting with a version with several limitations, most of which are later lifted.

\subsection{Basic Algorithm}

Our initial limitations / assumptions are as follows:

\begin{enumerate}
  \item $G$ contains no parentheses, i.e., given $G = (S, P, \NT, \T, \I)$ we require $\T \cap \{\verb|'('|, \verb|')'|\} = \emptyset$.

  \item $G'$ is constructed by adding parentheses to all non-terminals of $G$, i.e., $G' = (S, P', \NT, \T \cup \{\verb|'('|, \verb|')'|\}, \I \cup \NT)$ where $P' = P \cup \{ N: N -> \verb|'('|N\verb|')'| \mid N \in \NT \}$.

  \item No production has a right-hand side that matches a single non-terminal, i.e., $\forall (i:N->r) \in P.\ \NT \cap R(r) = \emptyset$\footnote{$R$ is here the more traditional language of a regular expression, i.e., $R(r) \subseteq (\T \cup \NT)^{*}$. It should be written more explicitly somewhere, but I'm lazy at the moment.}.
\end{enumerate}

We are looking for a counterexample to the resolvable property, i.e., a $t \in L(T)$ for which $\lnot \exists w' \in L(G').\ \parse(w') = \{t\}$. By the construction of $G'$, we have $t \in \parse(\yield(t))$, thus there is at least one (potentialy ambiguous) word $w'$ that can be parsed as $t$. Our task is thus to find a tree $t$ such that $\forall w' \in L(G').\ t \in \parse(w') -> \{t\} \subset \parse(w')$, i.e., a tree that can only be parsed with ambiguous words.

We begin by noting that parentheses cannot be added just anywhere; they must correspond to a node in the parse tree. For example, for the tree in Figure~\ref{fig:example-grammar:sem-tree} (which could be parsed from ''$1 + 2 * 3$'') it would be valid to add parentheses around ''$1 + 2$'' but not ''$2 * 3$''. As a consequence, adding parentheses restricts the possible parse trees, i.e., given two words $w'_1$ and $w'_2$ where the latter has added some parentheses we have $\parse(w'_1) \supseteq \parse(w'_2)$. We also note that adding double parentheses imposes no additional restriction, e.g., $\parse(\text{''}((1 + 2)) * 3\text{''})$ = $\parse(\text{''}(1 + 2) * 3\text{''})$.

There is thus a ''most restrictive word'' for every tree. If this word is ambiguous, then we have our counterexample\footnote{There is one more thing to note here to motivate the pushdown automata: the only form of ambiguity we have is one where there is another tree $t_2$ such that if $t \in \parse(w')$ then $t_2 \in \parse(w')$. I am convinced this is true for the basic case, and I believe it is still true for the remaining cases, but it needs to be proved. As such, I'm unsure how to motivate it here, because I think it'll require more complicated reasoning in later cases, and it'd be nice if there could be some commonality between the explanations.}.

We will now construct a pushdown automaton that recognizes these most restrictive words in such a way that there is a bijection between successful runs and trees in $L(T)$. As a running example, we will use the following (very simple) grammar $G$:

\setlength{\grammarindent}{3.5em}
\begin{grammar}
  <N> ::= succ: 's' <N>
    \alt zero: 'z'
\end{grammar}

\noindent We begin by constructing a DFA per production. This can be done in the standard way by constructing an NFA, then determinizing it, and optionally minimizing it.

\begin{center}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
    scale = 1,transform shape]

  \node[state,initial] (1) {$1$};
  \node[state] (2) [right of=1] {$2$};
  \node[state,accepting] (3) [right of=2] {$3$};
  \node[state,initial] (4) [below of=1, xshift=1cm] {$4$};
  \node[state,accepting] (5) [right of=4] {$5$};

  \path (1) edge              node {$s$} (2)
  (2) edge              node {$N$} (3)
  (4) edge              node {$z$} (5);

\end{tikzpicture}
\end{center}

We then combine them into a single pushdown automata by replacing each edge with a non-terminal label $p \xrightarrow{N} q$ with ($+\gamma$ means ''push $\gamma$'', while $-\gamma$ means ''pop $\gamma$''):

\begin{itemize}
  \item An edge $p \xrightarrow{'(', +(p, q)} p'$ for every initial state $p'$ in some DFA belonging to non-terminal $N$.
  \item An edge $q' \xrightarrow{')', -(p, q)} q$ for every final state $q'$ in some DFA belonging to non-terminal $N$.
\end{itemize}

\begin{center}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
    scale = 1,transform shape]

  \node[state,initial] (1) {$1$};
  \node[state] (2) [right of=1] {$2$};
  \node[state,accepting] (3) [right of=2] {$3$};
  \node[state,initial] (4) [below of=1, xshift=1cm] {$4$};
  \node[state,accepting] (5) [right of=4] {$5$};

  \draw
  (1) edge[above] node{$s$} (2)
  (4) edge[above] node{$z$} (5)
  (2) edge[bend left, below] node{$'(', +(2, 3)$} (1)
  (2) edge[bend left, left] node{$'(', +(2, 3)$} (4)
  (3) edge[loop above] node{$')', -(2, 3)$} (3)
  (5) edge[left, bend right] node{$')', -(2, 3)$} (3);

\end{tikzpicture}
\end{center}

\noindent Finally, we add a new initial state and a new final state, and connect them with the initial and final states belonging to the starting non-terminal:

\begin{center}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
    scale = 1,transform shape]

  \node[state] (1) {$1$};
  \node[state] (2) [right of=1] {$2$};
  \node[state] (3) [right of=2] {$3$};
  \node[state] (4) [below of=1, xshift=1cm] {$4$};
  \node[state] (5) [right of=4] {$5$};
  \node[state,initial] (s) [left of=4] {};
  \node[state,accepting] (e) [right of=5] {};

  \draw
  (s) edge[left] node{$'(', +s$} (1)
  (s) edge[above] node{$'(', +s$} (4)
  (3) edge[right] node{$')', -s$} (e)
  (5) edge[above] node{$')', -s$} (e)
  (1) edge[above] node{$s$} (2)
  (4) edge[above] node{$z$} (5)
  (2) edge[bend left, below] node{$'(', +(2, 3)$} (1)
  (2) edge[bend left, left] node{$'(', +(2, 3)$} (4)
  (3) edge[loop above] node{$')', -(2, 3)$} (3)
  (5) edge[left, bend right] node{$')', -(2, 3)$} (3);

\end{tikzpicture}
\end{center}

\noindent The resulting pushdown automaton has only a single source of non-determinism: the edges labelled \verb|'('|. Each one corresponds to one of the allowable child productions at that point in the parse tree.

%% \noindent Generalizing this to a complete grammar, we get the following property: a grammar $G$ is resolvable iff:

%% $$
%% \begin{array}{l}
%%   \forall t \in L(T).\\
%%   \quad \exists w' \in L(G').\\
%%   \qquad \forall t' \in L(T').\ \yield(t') = w' -> \unparen(t') = t
%% \end{array}
%% $$

@@

%% \begin{theorem}
%%   If $\{(, )\} \cap \T = \emptyset$, then any unambiguous word $w \in L(G)$ is resolvable.
%% \end{theorem}

%% \noindent \textbf{Proof.} Since $w$ is unambiguous in $G$ there is exactly one parse tree $t \in L(T)$ such that $\yield(t) = w$. Since $L(G) \subseteq L(G')$ we have that $w \in L(G')$. Furthermore, $w$ is unambiguous in $L(G')$ as well, since no added productions can apply (they all contain parentheses, which cannot be present in $w$). Additionally, since $L(T) \subseteq L(T')$, we have that $t \in L(T')$, and since $\yield(t) = w$, it must be the only parse tree for $w$ in $L(T')$. Finally, $\unparen(t) = t$.

%% %% Acknowledgments
%% \begin{acks}                            %% acks environment is optional
%%                                         %% contents suppressed with 'anonymous'
%%   %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
%%   %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
%%   %% acknowledge financial support and will be used by metadata
%%   %% extraction tools.
%%   This material is based upon work supported by the
%%   \grantsponsor{GS100000001}{National Science
%%     Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
%%   No.~\grantnum{GS100000001}{nnnnnnn} and Grant
%%   No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
%%   conclusions or recommendations expressed in this material are those
%%   of the author and do not necessarily reflect the views of the
%%   National Science Foundation.
%% \end{acks}

%% Bibliography
\bibliography{All}

%% %% Appendix
%% \appendix
%% \section{Appendix}

%% Text of appendix \ldots
